Stochastic Convolutional Sparse Coding

### 2018-11-12 cvpr2019/SCSCmodel.tex
fix grammar

\section{Stochastic Convolutional Sparse Coding}
\subsection{The Model}
\note{The previous description was not accurate: Eq.2 is not a variant
  of the CSC model, and does not replace Eq.1! Eq.2 is supposed to be
  used inside an outer loop to implement Eq.1. This outer loop is an
  implementation of CSC and Eg.1.  In order to really explain this
  properly, we need to introduce an outer loop parameter for both
  $\mask$ and $\mathcal{M}$.}

Based on the strong sparsity of the codes $\code$, we propose to
implement the CSC learning problem~\eqref{eq:CSCmodel} iteratively,
where at $\ite$-th iteration we only consider a random subset
$\mathcal{M}_k^{\ite}(\code_k)$ of the codes:
\begin{align}
    &\mini{\filter,\code}  \frac{1}{2}\|\signal - \sum_{k=1}^{K} \filter_k * \mathcal{M}_k^{\ite}(\code_k) \|_2^2 + \lambda \sum_{k=1}^{K}\| \mathcal{M}_k^{\ite}(\code_k) \|_1  \nonumber \\
    &\text{subject to} ~ \|\filter_k\|^2_2 \leq 1 ~~ \forall k \in \{1,\dots,K\}.
\end{align}
\todo{add outer loop variable}
\todo{There are some notational issues here -- convolution with an
  irregularly subsampled signal! Lets discuss on Sunday.}
Here $\mathcal{M}_k^{\ite}(\code_k)$ is the operation to subsample the
corresponding code $\code_k$ following a Bernoulli distribution with
probability $p$, drawn at current iteration.
%The probability value $p$ controls the percentage of
%the sparse codes that will be considered by the solver.
In the case of $p=1$, the proposed model is identical with the
classical CSC model. When $p<1$, the algorithm only solves a subset of
the sparse codes at chosen positions in each iteration, and
accordingly, the update of dictionary $d$ is based on the selected
portion of the sparse codes. Similar to solving the classical CSC
problem, we can apply a coordinate descent algorithm, alternating on
subproblems of $\mathcal{M}(\code)$ and $\filter$, to solve the above
bi-convex optimization problem. Specifically, the modified
minimization problem for $\code$ can be formulated as:
\begin{equation} \label{eq:updatingCode}
    \code^{\ite} = \minimize{\code} \frac{1}{2}\|\signal - (\Filter^{\ite-1} (\mask^{\ite})^\top) (\mask^{\ite} \code) \|_2^2 + \lambda \|\mask^{\ite} \code\|_1,
\end{equation}
\todo{add outer loop variable}
where $\Filter^{\ite-1} = [\Filter_1^{\ite-1}, \dots, \Filter_K^{\ite-1}] \in \mathbb{R}^{D
  \times DK}$, which is constructed from the dictionary learned by
$(\ite-1)$-th iteration, $\code = [\code_1, \dots, \code_K] \in
\mathbb{R}^{DK}$. The convolution operators are expressed as a matrix
multiplication so that $ \Filter \code = \sum_{k=1}^{K} \filter_k *
\code_k$. Therefore, each part of $\Filter$ is a Toeplitz
matrix. Matrix $\mask^{\ite}$ is a $\mathbb{R}^{pDk \times DK}$ random
diagonal matrix with $1$ for the entries corresponding to the sampled
sparse codes and $0$ elsewhere, performing the function of
$\mathcal{M}^{\ite}(\code)$. Matrix $\mask^{\ite}$ will be redrawn at each
iteration. It is worth stressing that $(\Filter^{\ite-1} (\mask^{\ite})^\top)$ is not
performing a subsampling operation on the dictionary, \note{not sure
  what this is supposed to say:} which aims to remove the columns in $\Filter^{\ite}$ those are filtered out
by $\mathcal{M}^{\ite}(\code)$.
\note{See notation issue above; we need to discuss this next week:}
Due to the introduced subsampling matrix, the
convolution operator cannot be implemented in the Fourier
domain. However, due to the subsampling
strategy, the number of variables that need to be computed for this
subproblem is $pDK$ instead of $DK$, which leads to a reduction
of the spatial-domain computation time by a factor of $\frac{1}{p}$.

After computing the subsampled sparse codes $\mathcal{M}^{\ite}(\code)$, we can then project them
onto the original spatial support by substituting $0$ for the
masked-out values, obtaining $z^{\ite}$. Afterwards, the dictionary can be updated by:
\begin{equation} \label{eq:updatingFilter}
\begin{split}
   & \filter^{\ite} = \minimize{\filter} \frac{1}{2}\|\signal - \Code^{\ite} \filter \|_2^2 \\
   & \text{subject to}  ~ \|\filter_k\|^2_2 \leq 1 ~ \forall k \in \{1,\dots,K\},
\end{split}
\end{equation}
\todo{add outer loop variable} where $\Code^{\ite}= [\Code_1^{\ite}, \dots, \Code_K^{\ite}]
\in \mathbb{R}^{D \times MK}$ is a concatenation of Toeplitz matrices,
and $\Code_k^{\ite}$ is constructed from the associated $\code_k^{\ite}$, $\filter=
[\filter_1,\dots,\filter_K] \in \mathbb{R}^{MK}$ such that $ \Code
\filter = \sum_{k=1}^{K} \filter_k * \code_k$. In typical CSC
settings, $M \ll D$, hence there is no need to perform subsampling on
the dictionary.

@@

\begin{algorithm}[H]
\caption{SOCSC} \label{algo:SOCSC}
\begin{algorithmic}[1]
\State $\text{Initialize} ~ t=0, ~\filter^t, ~p, ~\surC^t = 0, ~\surB^t = 0$
\While {not converge}
    \State $t \gets t+1$
    \State $ \text{draw } \signal^t \text{ from training images} $
    \State $ \text{Randomly sample }\code^t \text{ with rate } p $
    \State $ \text{Compute } \code^t \text{ by solving problem~(\ref{eq:updatingCodeOnline}) using } \filter^{t-1}$
    \State Compute $\surC^t$ and $\surB^t$ by Eq.~\ref{eq:updateSur}
    \State Compute $\filter^t$ by solving problem~(\ref{eq:updatingFilterOnline})
\EndWhile
\end{algorithmic}
\todo{Find the best place for this algorithm (probably higher up)}
\end{algorithm}

In order to address this scalability issue, we can tackle the
Stochastic CSC problem in an online fashion. In the online learning
setting as shown in Algorithm~\ref{algo:SOCSC}, each iteration only
draws one or a subset (mini-batch) of the total training images, hence
the complexity per loop is independent of the training sample
size. Then, given the sampled image $\signal^t$ at $t$-th iteration,
we can compute the corresponding subsampled sparse codes $\code^t$ by
\begin{equation} \label{eq:updatingCodeOnline}
    \code^t = \minimize{\code} \frac{1}{2}\|\signal^t - (\Filter^{t-1} (\mask^t)^\top)(\mask^t \code) \|_2^2 + \lambda \|\mask^t \code\|_1.
\end{equation}
\note{Here we have $t$ as an iteration variable now -- I don't think
  this is a good choice ($t$ is usually reserved for continuous
  values, better would be $i$ or $j$, and we need to introduce these
  earlier in the paper.}
The only difference between Eq. ~\ref{eq:updatingCode} is that $\signal^t$ only contains a portion of the total images. After obtaining
the sparse codes, the dictionary is updated by:
\begin{equation}
\begin{split}
    &\filter^t = \minimize{\filter} \frac{1}{2t}\sum_{i=1}^{t} \|\signal^i - \Code^i \filter \|_2^2 \\
    &\text{subject to} \quad \|\filter_k\|^2_2 \leq 1 ~ \forall k \in \{1,\dots,K\}.
\end{split}
\end{equation}
Note that updating the dictionary in this fashion involves all of the
past training images and sparse codes. Using techniques developed for
regular (non-convolutional) dictionary
learning~\cite{mairal2009online,mairal2010online}, we can get rid of
explicitly storing this data by introducing two surrogate matrices
$\surC \in \mathbb{R}^{KM \times KM}$ and $\surB \in \mathbb{R}^{KM
  \times 1}$. These carry all of the required information for updating
$\filter$, and can be iteratively updated by:
\begin{equation} \label{eq:updateSur}
\begin{split}
    \surC^t  = \frac{t-1}{t} \surC^{t-1} + \frac{1}{t}(\Code^t)^\top \Code^t \\
    \surB^t  = \frac{t-1}{t} \surB^{t-1} + \frac{1}{t}(\Code^t)^\top \signal^t
\end{split}
\end{equation}
With these surrogate matrices, the updated dictionary can be obtained
by solving
\begin{equation} \label{eq:updatingFilterOnline}
\begin{split}
    & \filter^t = \minimize{\filter} \frac{1}{2} \filter^\top \surC^t\filter - \filter^\top \surB^t \\& \text{subject to} \quad \|\filter_k\|_2^2 \leq 1 ~ \forall k \in \{1,\dots,K\}.
\end{split}
\end{equation}
Problem~\eqref{eq:updatingCodeOnline} and
problem~\eqref{eq:updatingFilterOnline} are solved in the same way as
that for SBCSC.

>>>>

\section{Stochastic Convolutional Sparse Coding}
\subsection{The Model}
Based on the strong sparsity of the codes $\code$, we propose to
implement the CSC learning problem~\eqref{eq:CSCmodel} iteratively,
where at $\ite$-th iteration we only consider a random subset
$\mathcal{M}_k^{\ite}(\code_k)$ of the codes:
\begin{equation} \label{eq:updatingFilter}
\begin{split}
    & (\filter^{\ite},\code^{\ite}) =  \\
    &\minimize{\filter,\code}  \frac{1}{2}\|\signal - \sum_{k=1}^{K} \filter_k * \mathcal{M}_k^{\ite}(\code_k) \|_2^2 + \lambda \sum_{k=1}^{K}\| \mathcal{M}_k^{\ite}(\code_k) \|_1  \\
    &\text{subject to} ~ \|\filter_k\|^2_2 \leq 1 ~~ \forall k \in \{1,\dots,K\}.
\end{split}
\end{equation}
Here $\mathcal{M}_k^{\ite}(\code_k)$ is a $\mathbb{R}^D \to \mathbb{R}^D$ operator subsampling the
corresponding code $\code_k$ following a Bernoulli distribution with
probability $p$, drawn at current iteration.
%The probability value $p$ controls the percentage of
%the sparse codes that will be considered by the solver.
In the case of $p=1$, the proposed model is identical with the
classical CSC model. When $p<1$, the algorithm only solves a subset of
the sparse codes at chosen positions in each iteration, and
accordingly, the update of dictionary $d$ is based on the selected
portion of the sparse codes. Similar to solving the classical CSC
problem, we can apply a coordinate descent algorithm, alternating on
subproblems of $\mathcal{M}^{\ite}(\code)$ and $\filter$, to solve the above
bi-convex optimization problem. Specifically, the modified
minimization problem for $\code$ can be formulated as:
\begin{equation} \label{eq:updatingCode}
    \code^{\ite} = \minimize{\code} \frac{1}{2}\|\signal - \Filter^{\ite-1} \mask^{\ite} \code \|_2^2 + \lambda \|\mask^{\ite} \code\|_1,
\end{equation}
where $\Filter^{\ite-1} = [\Filter_1^{\ite-1}, \dots, \Filter_K^{\ite-1}] \in \mathbb{R}^{D
  \times DK}$, which is constructed from the dictionary learned by
$(\ite-1)$-th iteration, $\code = [\code_1, \dots, \code_K] \in
\mathbb{R}^{DK}$. The convolution operators are expressed as a matrix
multiplication so that $ \Filter \code = \sum_{k=1}^{K} \filter_k *
\code_k$. Therefore, each part of $\Filter$ is a Toeplitz
matrix. Matrix $\mask^{\ite}$ is a $\mathbb{R}^{Dk \times DK}$ random binary
diagonal matrix with $1$s on the diagonal entries corresponding to the sampled
sparse codes and $0$ elsewhere, performing the function of
$\mathcal{M}^{\ite}(\code)$, which satisfies $\mathrm{Tr}(\mask^{\ite}) = pDK$. When $p=1$, $\mask^{\ite}$ is just an identity matrix.
Due to the introduced subsampling matrix, the
convolution operator cannot be implemented in the Fourier
domain. However, owing to the subsampling
strategy, the number of variables that need to be computed for this
subproblem is $pDK$ instead of $DK$, which leads to a reduction
of the spatial-domain computation time by a factor of $\frac{1}{p}$.

After computing the subsampled sparse codes $\mathcal{M}^{\ite}(\code)$, we can then project them
onto the original spatial support by substituting $0$ for the
masked-out values, obtaining $\code^{\ite}$. Afterwards, the dictionary can be updated by:
\begin{equation} \label{eq:updatingFilter}
\begin{split}
   & \filter^{\ite} = \minimize{\filter} \frac{1}{2}\|\signal - \Code^{\ite} \filter \|_2^2 \\
   & \text{subject to}  ~ \|\filter_k\|^2_2 \leq 1 ~ \forall k \in \{1,\dots,K\},
\end{split}
\end{equation}
where $\Code^{\ite}= [\Code_1^{\ite}, \dots, \Code_K^{\ite}]
\in \mathbb{R}^{D \times MK}$ is a concatenation of Toeplitz matrices,
and $\Code_k^{\ite}$ is constructed from the associated $\code_k^{\ite}$, $\filter=
[\filter_1,\dots,\filter_K] \in \mathbb{R}^{MK}$ such that $ \Code
\filter = \sum_{k=1}^{K} \filter_k * \code_k$. In typical CSC
settings, $M \ll D$, hence there is no need to perform subsampling on
the dictionary.

@@

\begin{algorithm}[H]
\caption{SOCSC} \label{algo:SOCSC}
\begin{algorithmic}[1]
\State $\text{Initialize} ~ t=0, ~\filter^t, ~p, ~\surC^t = 0, ~\surB^t = 0$
\While {not converge}
    \State $t \gets t+1$
    \State $ \text{draw } \signal^t \text{ from training images} $
    \State $ \text{Randomly sample }\code^t \text{ with rate } p $
    \State $ \text{Compute } \code^t \text{ by solving problem~(\ref{eq:updatingCodeOnline}) using } \filter^{t-1}$
    \State Compute $\surC^t$ and $\surB^t$ by Eq.~\ref{eq:updateSur}
    \State Compute $\filter^t$ by solving problem~(\ref{eq:updatingFilterOnline})
\EndWhile
\end{algorithmic}
\end{algorithm}

In order to address this scalability issue, we can tackle the
Stochastic CSC problem in an online fashion. In the online learning
setting as shown in Algorithm~\ref{algo:SOCSC}, each iteration only
draws one or a subset (mini-batch) of the total training images, hence
the complexity per loop is independent of the training sample
size. Then, given the sampled image $\signal^t$ at $t$-th iteration,
we can compute the corresponding subsampled sparse codes $\code^t$ by
\begin{equation} \label{eq:updatingCodeOnline}
    \code^t = \minimize{\code} \frac{1}{2}\|\signal^t - \Filter^{t-1}\mask^t \code \|_2^2 + \lambda \|\mask^t \code\|_1.
\end{equation}
The only difference between Eq. ~\ref{eq:updatingCode} is that $\signal^t$ only contains a portion of the total images. After obtaining
the sparse codes, the dictionary is updated by:
\begin{equation}
\begin{split}
    &\filter^t = \minimize{\filter} \frac{1}{2t}\sum_{i=1}^{t} \|\signal^i - \Code^i \filter \|_2^2 \\
    &\text{subject to} \quad \|\filter_k\|^2_2 \leq 1 ~ \forall k \in \{1,\dots,K\}.
\end{split}
\end{equation}
Note that updating the dictionary in this fashion involves all of the
past training images and sparse codes. Using techniques developed for
regular (non-convolutional) dictionary
learning~\cite{mairal2009online,mairal2010online}, we can get rid of
explicitly storing this data by introducing two surrogate matrices
$\surC \in \mathbb{R}^{KM \times KM}$ and $\surB \in \mathbb{R}^{KM
  \times 1}$. These carry all of the required information for updating
$\filter$, and can be iteratively updated by:
\begin{equation} \label{eq:updateSur}
\begin{split}
    \surC^t  = \frac{t-1}{t} \surC^{t-1} + \frac{1}{t}(\Code^t)^\top \Code^t \\
    \surB^t  = \frac{t-1}{t} \surB^{t-1} + \frac{1}{t}(\Code^t)^\top \signal^t
\end{split}
\end{equation}
With these surrogate matrices, the updated dictionary can be obtained
by solving
\begin{equation} \label{eq:updatingFilterOnline}
\begin{split}
    & \filter^t = \minimize{\filter} \frac{1}{2} \filter^\top \surC^t\filter - \filter^\top \surB^t \\& \text{subject to} \quad \|\filter_k\|_2^2 \leq 1 ~ \forall k \in \{1,\dots,K\}.
\end{split}
\end{equation}
Problem~\eqref{eq:updatingCodeOnline} and
problem~\eqref{eq:updatingFilterOnline} are solved in the same way as
that for SBCSC.
